---
title: "Asymmetric Markov Chain Tutorial"
author: "Stephen Bates and Wenshuo Wang"
date: "`r Sys.Date()`"
---

This notebook demonstrates the knockoff construction for the asymmetric Markov Chain experiments. The data is generated as follows. Let
$$
Z_j\stackrel{i.i.d.}{\sim} \frac{I\cdot \mid  Y_\text{G} \mid  -(1-I)\cdot Y_\text{E}- \mu}{\sigma} \ \text{ for } j=1,\dots,p,
$$
where $\mu$ and $\sigma$ are chosen so that the variables are centered with unit variance, and then define
$$
X_1=Z_1, \quad X_{j+1}=\rho_j X_j + \sqrt{1-\rho_j^2}Z_{j+1} \ \text{ for } j=2,\dots,p.
$$
Section 5.2.2 of the accompanying paper presents simulation results in this setting.

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
#library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```


## Multiple-try Metropolis

We first demonstrate the multiple-try metropolis technique (Section 3.3 of the paper) with the recommended settings.

```{r}
# simulation parameters
p = 50 # dimension of the random vector X
numsamples = 100 # number of samples to generate knockoffs for
rhos = rep(0.6,p-1) # the correlations

#algorithm parameters
halfnumtry = 4 # m/half number of candidates
stepsize = 1.5 # step size in the unit of 1/\sqrt((\Sigma)^{-1}_{jj})

#mean/variance constants
asymean = 1/sqrt(2*pi)-1/2
variance = 1.5-(1/sqrt(2*pi)-1/2)^2
```

```{r}
library(foreach)
source("../asymmetric/asym_MTM_core.R") #load the knockoff functions
```

First, we compute the stepsize of the grid of proposals for each variable. The recommended stepsize for variable $j$ (Section 3.3 of the paper) is 
$$
1.5 / \sqrt{(\Sigma^{-1})_{jj}}.
$$

```{r}
#compute the proposal size at each step
#using an exact analytic calculation
quantile.x = matrix(0,ncol=2*halfnumtry+1,nrow=p)
sds = rep(0,p)
sds[1] = sqrt(1-rhos[1]^2)
for(i in 2:(p-1)){
  sds[i] = sqrt((1-rhos[i-1]^2)*(1-rhos[i]^2)/(1-rhos[i-1]^2*rhos[i]^2))
}
sds[p] = sqrt(1-rhos[p-1]^2)
for(i in 1:p){
  quantile.x[i,] = ((-halfnumtry):halfnumtry)*sds[i]*stepsize
}
```

Then, we sample the Markov chain and the knockoffs with the `SCEP.MH.MC` function.

```{r warning=FALSE}
# sample the variable and generate the knockoffs
temp = matrix(0,ncol=2*p,nrow=1)
start = Sys.time()
result_mtm = foreach(i=1:numsamples,.combine='rbind') %do% {
  #sample the MC
  if(runif(1)>=0.5){
    temp[1] = (abs(rnorm(1))-asymean)/sqrt(variance)
  } else{
    temp[1] = (-rexp(1)-asymean)/sqrt(variance)
  }
  for(j in 2:p){
    if(runif(1)>=0.5){
      temp[j] = ((abs(rnorm(1))-asymean)/sqrt(variance))*sqrt(1-rhos[j-1]^2) + rhos[j-1]*temp[j-1]
    } else{
      temp[j] = ((-rexp(1)-asymean)/sqrt(variance))*sqrt(1-rhos[j-1]^2) + rhos[j-1]*temp[j-1]
    }
  }
  
  #generate the knockoffs
  temp[(p+1):(2*p)] = SCEP.MH.MC(temp[1:p],0.999,quantile.x)
  temp
}
end = Sys.time()
print(paste0("Average time per observation + knockoff (seconds): ", round((end - start) / numsamples, digits = 3)))

dim(result_mtm)
```

We can compute the mean correlation between $X_j$ and $\tilde{X}_j$ to measure the knockoff quality:

```{r}
cors = c()
for(j in 1:p) {
  cors = c(cors, cor(result_mtm[, j], result_mtm[, j+p]))
}
mean(cors)
```

## Covariance-guided proposals

We can also generate knockoffs with the covariance-guided proposals (Section 3.2 of the paper) with the `SCEP.MH.MC.COV` function. These proposals are motivated by a Gaussian approximation (indeed, they are optimal for Gaussian distributions), but they produce *exact* knockoffs because of the Metropolis--Hastings correction.

```{r}
set.seed(100)

temp = matrix(0,ncol=2*p,nrow=1)
start = Sys.time()
result_cov = foreach(i=1:numsamples,.combine='rbind',.packages = 'MASS') %do% {
  #sample the MC
  if(runif(1)>=0.5){
    temp[1] = (abs(rnorm(1))-asymean)/sqrt(variance)
  } else{
    temp[1] = (-rexp(1)-asymean)/sqrt(variance)
  }
  for(j in 2:p){
    if(runif(1)>=0.5){
      temp[j] = ((abs(rnorm(1))-asymean)/sqrt(variance))*sqrt(1-rhos[j-1]^2) + rhos[j-1]*temp[j-1]
    } else{
      temp[j] = ((-rexp(1)-asymean)/sqrt(variance))*sqrt(1-rhos[j-1]^2) + rhos[j-1]*temp[j-1]
    }
  }

  # generate the knockoffs
  xk = SCEP.MH.MC.COV(temp[1:p],1,rep(0,p), prop.mat, cond.means.coeff, cond.vars)
  temp[(p+1):(2*p)] = xk[1:p]
  temp
}
end = Sys.time()
print(paste0("Average time per observation + knockoff (seconds): ", round((end - start) / numsamples, digits = 3)))

dim(result_cov)
```

We can again compute the mean correlation between $X_j$ and $\tilde{X}_j$ to check the knockoff quality:

```{r}
cors = c()
for(j in 1:p) {
  cors = c(cors, cor(result_cov[, j], result_cov[, j+p]))
}
mean(cors)
```
