---
title: "Ising Model Knockoffs Tutorial"
author: "Stephen Bates and Wenshuo Wang"
date: "`r Sys.Date()`"
---

```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
#library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=TRUE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```

This notebook illustrates the usage of the knockoff construction scripts to create knockoffs for variables that are distributed according to the Ising model on a grid:
$$
P(X) = \frac{1}{Z(\beta_0)} e^{\beta_0 \sum_{(i,i^\prime) \in E} X_i X_{i^\prime}}
$$
where $i$ and $i^\prime$ are each coordinates in the $d \times d$ grid, $E$ is the set of edges of the grid, and each $X_i \in \{-1, 1\}$.



## Knockoffs for the Ising model

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(reshape2)
library(knockoff)
library(bayess)
library(latex2exp)
source("../Ising/Ising.r")
```

```{r}
# Simulation #parameters
n = 500 #number of samples
d = 5 #5x5 grid
beta_0 = .5 #interaction strength
```

We create `r n` samples from a `r d`x`r d` Ising model using the `bayess` package.

```{r}
#sample from the Ising model
set.seed(100)
X = t(replicate(n, as.vector(isinghm(niter = 50, n = d, m = d, beta = 2*beta_0))))
dim(X)
```

Next, we generate knockoffs for each sample using the `ising_knockoff` function.

```{r}
start = Sys.time()
set.seed(100)
Xk = c()
for(i in 1:n) {
  x = matrix(X[i, ], nrow = d, ncol = d)
  xk = ising_knockoff(x, d, d, beta_0)
  Xk = rbind(Xk, as.vector(xk))
}
end = Sys.time()
print(paste0("Average time per knockoff (seconds): ", (end - start) / n))

dim(Xk)
```

The mean absolute correlation between each variable and it's corresponding knockoff is

```{r}
cors = c()
for(j in 1:(d*d)) {
  cors = c(cors, cor(X[, j], Xk[, j]))
}
mean(cors)
```

### Quality across coordinates.

Next, we investigate the quality of the knockoffs at each location in the grid. We compute the correlation between $X_{i,j}$ and $\tilde{X}_{i,j}$ for all each index  $(i,j)$ in the grid and plot the resulting data as a heatmap.

```{r}
#compute correlation between original variable and knockoff at each coordinate.
cors = data.frame(stringsAsFactors = FALSE)
for(i in 1:d) {
  for(j in 1:d) {
    cors = rbind(cors, c("knockoff", i, j, 
                         cor(X[, i + (j - 1) * d], Xk[, i + (j - 1) * d])), 
                 stringsAsFactors = FALSE)
  }
}
colnames(cors) = c("type", "i", "j", "ko_orig_corr")
cors$ko_orig_corr = as.numeric(cors$ko_orig_corr)
cors$i = as.numeric(cors$i)
cors$j = as.numeric(cors$j)
cors = cors %>% 
  mutate(order = d * (i - 1) + j)
```

```{r}
#plot the knockoff quality at each coordinate
quality_grid = ggplot(cors, aes(x = i, y = j, fill = ko_orig_corr)) +
  geom_tile(color = "white") +
  scale_fill_continuous(high = "#000000", low = "#56B1F7", limits = c(.3, .8)) + 
  scale_y_reverse() + 
  coord_equal() +
  labs(fill = "cor") +
  theme_bw() + 
  theme(text = element_text(size = 10)) +
  xlab("j") + 
  ylab("k") + 
labs(title = TeX("$5 \\times 5$ grid, $\\beta_0 = 0.5$")) +
  coord_fixed() + 
  theme(plot.title = element_text(size=10)) +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme(legend.position = "bottom") +  
  theme(legend.text=element_text(size=8))

quality_grid
```

We see that the variables on the edges of the grid are less correlated with their knockoffs, as expected.

## Knockoffs for large grids

For large grids, we can still generate knockoffs efficiently using the divid-and-conquer technique, which is implemented in the `ising_knockoff_slice` function. We demonstrate this technique here.

First, we sample from the Ising model on a grid.

```{r}
# Simulation #parameters
n = 100 #number of samples
d = 20 #20x20 grid
beta_0 = .5 #interaction strength
width = 3 #divide-and-conquer width
```

```{r}
#sample from the ising model
set.seed(100)
X = t(replicate(n, as.vector(isinghm(niter = 50, n = d, m = d, beta = 2*beta_0))))
dim(X)
```

Next, we sample the knockoffs using the divide-and-conquer technique.

```{r}
set.seed(100)
start = Sys.time()
Xk = c()
for(i in 1:n) {
  x = matrix(X[i, ], nrow = d, ncol = d)
  xk = ising_knockoff_slice(x, d, d, beta_0, max_width = width)
  Xk = rbind(Xk, as.vector(xk))
}
end = Sys.time()
print(paste0("Average time per knockoff (seconds): ", (end - start) / n))

dim(Xk)
```

We again compute the mean correlation between $X_{j,k}$ and $\tilde{X}_{j,k}$:

```{r}
cors = c()
for(j in 1:(d*d)) {
  cors = c(cors, cor(X[, j], Xk[, j]))
}
mean(cors)
```
